Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging

### 2013-02-27 HRUIonGPU/jrnl_article.tex
Done some writing

To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
In time we have that
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculate from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ (cite JOE?). However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$. Note that sliding will break independence along the sliding dimension, thus, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is increased to the maximum number of threads per block (currently 1024) with this approach. However, for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.

@@

\subsection{Solving Multiple Small Linear Systems}
The literature has mostly focused on large systems when it comes to solving systems of linear equations on the GPU. GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems. The reason for this is two-fold. First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU. Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case. It is therefore not given that solving on the GPU is faster than solving on the CPU in general. However, there are two arguments for solving on the GPU. First the CPU, which might already run a lot of ultrasound related algorithms, is offloaded. Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to GPU, is avoided.

For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels in one kernel launch. The solver has been made available online for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices. All solvers have been benchmarked against the latest Intel MKL libraries.

@@

\subsection{Compute Beamformer Output}
To calculate the beamformer output, equation (\ref{eq:z_mv}) has been implemented using one thread per element-weight ($L$ threads per weight vector). After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$. 
%Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$.

The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increasing the weight vector to length $M$. Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate 
\begin{align}\label{eq:subsum}
\breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
\end{align}
Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory. %The last sum can also be performed using a binary reduction sum across several threads for increased speed.

@@

For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of beams contains almost all received energy. Nilsen et al. concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space. Hence, for  $\lambda /2$ spaced arrays the inversion steps is found to be constant with respect to the number of array elements.

@@

Another option is to transform the element space covariance matrix to beamspace. 
\begin{align}
\mat{\hat{R}}_\text{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H.
\end{align}
The advantages of using this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space. In this paper we will focus on the first approach, and how it has been implemented on the GPU.
%By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$. Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space. 
In both cases solving has to be performed against a beamspace steering vector
\begin{align}
\vec{a}_\text{BS} &= \mat{B}\vec{1}_L = \sqrt{L}\vec{e}_1,
\end{align}
where $\vec{e}_i$ has the value one in the i$^{th}$ position and zeroes elsewhere.

@@

Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap across subarrays, but with an increase in task level granularity (as discussed in Section \ref{sec:calcR}). In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio ($M/N_b$). Using one thread per beam is therefore not feasible for small $N_b$ large $M$ configurations. The solution is to combine SBS with full computations. A number of thread groups is created that targets different sets of subarrays which they transform using the SBS algorithm. In that way, the number of threads per sample is increased and thus the memory per thread ratio is decreased.

Transforming subarrays into beamspace yields subarrays that no longer overlap. This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available. On the other hand, since $N_b$ should be small ($N_b < 32$), one thread per element in $\mat{\hat{R}}_\text{BS}$ is now possible. A kernel similar to the one described in \cite{Chen2011} has therefore been implemented. In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high occupancy of the GPU.

>>>>

To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
In time we have that
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculate from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ (cite JOE?). However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$ (if $L=M/2$). Note that sliding will break independence along the sliding dimension, thus, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is increased to the maximum number of threads per block (currently 1024) with this approach. However, for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.

@@

\subsection{Solving Multiple Small Linear Systems}
The literature has mostly focused on large systems when solving systems of linear equations on the GPU. GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems. The reason for this is two-fold. First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU. Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case. It is therefore not given that solving on the GPU is faster than solving on the CPU in general. However, there are two arguments for solving on the GPU. First the CPU, which might already run a lot of ultrasound related algorithms, is offloaded. Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to GPU, is avoided.

For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels in one kernel launch. The solver is available online for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices. All solvers have been benchmarked against the latest Intel MKL libraries.

@@

\subsection{Compute Beamformer Output}
To calculate the beamformer output, equation (\ref{eq:z_mv}) has been implemented using one thread per element-weight ($L$ threads per weight vector). After the solution vector $\vec{b}$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}$. 
%Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$.

The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increasing the weight vector to length $M$. Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate 
\begin{align}\label{eq:subsum}
\breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
\end{align}
Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to  GPU DRAM. %The last sum can also be performed using a binary reduction sum across several threads for increased speed.

@@

For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of beams contains almost all received energy. Nilsen et al. concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space. Hence, for $\lambda$-factor spaced arrays the inversion steps is found to be constant with respect to the number of array elements.

@@

Another option is to transform the element space covariance matrix to beamspace. 
\begin{align}
\mat{\hat{R}}_\text{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H.
\end{align}
The advantages of this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space. In this paper we will focus on the first approach, and how it has been implemented on the GPU.
%By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$. Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space. 
In both cases solving has to be performed against a beamspace steering vector
\begin{align}
\vec{a}_\text{BS} &= \mat{B}\vec{1}_L = \sqrt{L}\vec{e}_1,
\end{align}
where $\vec{e}_i$ has the value one in the i$^{th}$ position and zeroes elsewhere.

@@

Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap between subarrays, but with an increase in task level granularity (as discussed in Section \ref{sec:calcR}). In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio ($M/N_b$). Using one thread per beam is therefore not feasible for small $N_b$ large $M$ configurations. The solution is to combine SBS with full computations. A number of thread groups is created that targets different sets of subarrays which they transform using the SBS algorithm. In that way, the number of threads per sample is increased and thus the memory per thread ratio is decreased.

Transforming subarrays into beamspace yields subarrays that no longer overlap. This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available. On the other hand, since $N_b$ should be small ($N_b < 32$), one thread per element in $\mat{\hat{R}}_\text{BS}$ is now an option. A kernel similar to the one described in \cite{Chen2011} has therefore been implemented. In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high occupancy of the GPU.
